On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using Parallel Global Search Algorithm

### 2022-10-11 2022/PAVT Special Issue/template.tex

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The article is devoted to the problem of developing a kinetic model of a complex chemical reaction using a parallel optimization method. The design of the kinetic model consists in finding the kinetic parameters of the reaction, which cannot be calculated analytically, and since the chemical reaction involves many stages, the optimization problem is multi-extreme. As a chemical reaction, the process of catalytic isomerization of the pentane-hexane fraction is considered, which is now important due to the switch of the oil refining industry to the production of gasoline corresponding to the Euro-5 standard. To solve the optimization problem, the authors have developed parallel global search algorithm and a program based on Lipschitz optimization. The kinetic parameters found made it possible to develop a mathematical model of the process, which gives good agreement with industrial data.}

@@

In the complex problems of chemical kinetics, kinetic constants of reactions cannot be found analytically, they can only be found using numerical methods (see, for example, \cite{Zaynullin2020,Enikeeva2020,Koledina2019}). In this case, the quality criterion (objective function) is set in the form of a certain algorithm, which includes the stage of numerical modeling, which, in turn, makes it compute-intensive. Furthermore, in the inverse problems of chemical kinetics, the objective function can be substantially multiextremal, i.e., having many local extremes along the global one.

@@

It is also possible to construct deterministic methods of global search, which are different from brute force algorithms and metaheuristic algorithms. Such methods are associated with the presence and consideration of a priori assumptions about the problem properties. 
One of the natural assumptions about the problem is that changing the values of an objective function is limited by changes of its argument. In this case, the function satisfies the Lipschitz condition, and the problem itself is called the Lipschitz global optimization problem. The central direction of the development of modern methods of Lipschitz optimization is the development of parallel algorithms. It is assumed that the values of the objective function are determined as a result of solving some other task and require a costly computational experiment.

@@

The mathematical model involves such parameters as the raw material flow rate $F$ in the reactor and temperature $T$, but these two parameters also depend on $\tau$, therefore it is necessary to write equations on them, because technological process is non-isothermal:
\begin{linenomath}
\begin{equation}\label{eq_temper}
\dfrac{dT}{d\tau} = \dfrac{\sum_{i=1}^I\left( \dfrac{dx_i}{d\tau} \cdot \Delta H^i_{f,T} \right) }{\sum_i^J \left(x_i C^T_{p_i} \right) },
\end{equation}\label{eq_flow}
\begin{equation}
\dfrac{dF}{d\tau} = \sum_{i=1}^I  \dfrac{x_i}{d\tau}
\end{equation}
\end{linenomath}
with initial condition: at $\tau = 0$, $T(0) = T^0$, where $T$ is temperature, K, $\Delta H^i_{f,T}$ is enthalpy of the $i$-th component formation at temperature $T$, J/mol, $C^T_{p_i}$ is specific heat capacity of the $i$-th component at temperature $T$, (J/mol$\cdot$K), $F$ is mole discharge of the flow, kmol/h.

@@

\begin{linenomath}
\begin{equation}
FF = \dfrac{\sum_{i=1}^I \left| \dfrac{y_i^{exp} - y_i^{calc}}{y_i^{exp}} \right|}{I} \rightarrow \min,
\end{equation}
\end{linenomath}

where I is the number of observed parameters (I=19, of which 17 parameters are $x_i$, one parameter is $T$ and one parameter is $F$),  $y_{i}^{exp}$ are industrual data, $y_{i}^{calc}$ are numerical values of concentrations of reaction substances obtained by solving a direct kinetic problem. As a result of solving the inverse problem, the kinetic parameters $k^0_j$ and $E_j$ included in the Arrhenius equation \eqref{eq_rate} will be obtained. Since the number of reaction stages is 48, the kinetic model will contain 96 kinetic parameters (activation energy $E_j$ and pre-exponential multiplier $k^0_j$ for each stage), however, the authors decided to fix some of the parameters as already known for chemical reasons, so the number of optimized parameters was reduced to 24.

@@

As stated above, from a mathematical point of view, the inverse problem of chemical kinetics can be considered a Lipschitz global optimization problem. 
In general, the problem of the specified class can be formulated as follows:
\begin{gather}
 \phi^* = \phi(y^\ast)=\min{\left\{\phi(y):y\in D\right\}}, \label{problemN}\\
 D=\left\{y\in R^N: a_i\leq y_i \leq b_i, \;  1\leq i \leq N\right\} \label{D},
\end{gather}
where $a,b \in R^N$ are boundaries of the search domain, and the objective function $\phi(y)$ satisfies the Lipschitz condition
\begin{equation}\label{Lip}
\left|\phi(y_1)-\phi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D.
\end{equation}

The criterion $\phi(y)$ is assumed to be multiextremal and is given in the form of some computational algorithm. The input of this algorithm is supplied with a parameter vector, and the output is the corresponding function value. Such functions are often referred to as ``black-box'' functions. It is also assumed that each calculation of the function value at the search domain point (hereinafter also referred to as trial) is a laborious operation and requires significant computational resources. As noted in the introduction, this statement of the problem fully corresponds to the inverse problem of chemical kinetics.

@@

When describing a parallel algorithm, let us assume that there is one process master and $p$ workers. The upper index will indicate the serial number of the trial performed, and the lower index will be used to number the trials in ascending order of the coordinate.

@@

Step 1. Renumber he point of the set 
\[
X_k = \left\{x^1, x^2,...,x^{k+p} \right\},
\]
which contains all the points at which the trials either have been carried out or are being carried out by the lower index in ascending order, i.e.
\[
0=x_1<x_2<...<x_{x+p}=1.
\]
Step 2. Calculate values
\[
M_1=\max \left\{ \frac{ \left|z_i - z_{i-1} \right|}{(x_i-x_{i-1})^{1/N}} : x_{i-1} \notin I_k, x_i \notin I_k, 2\leq i\leq k+p \right\},
\]
\[
M_2=\max \left\{ \frac{ \left|z_{i+1} - z_{i-1} \right|}{(x_{i+1}-x_{i-1})^{1/N}} : x_i \in I_k, 2\leq i < k+p \right\},
\]
\[
M=\max\{M_1,M_2\},
\]
where $z_i=\phi(y(x_i))$, if $x_i \notin I_k, \; 1\leq i \leq k+p$.  If the value $M$ is 0, assign $M=1$. The values $z_i=\phi(y(x_i))$ at $x_i \in I_k$ are undefined because the trials at $x_i \in I_k$ are not finished yet.

@@

The master stops the algorithm if either the condition $\Delta_{t}<\varepsilon$ (where $t$ is number of the interval with the maximum characteristic from Step 4) or the condition $k>K_{max}$ (where $k$ is the number of trials performed) is met.
The real number $\varepsilon>0$ and the integer $K_{max} \geq 1$ are the algorithm parameters which correspond to the accuracy of the solution search and the maximum number of trials.

The sequential global search algorithm, on the basis of which the proposed asynchronous algorithm was implemented, allows the following interpretation \cite{Molinaro2001}. Using the points of previously performed trials, it is possible to construct a minorant of the objective function, for which the value $-R(i)$, where $R(i)$ is the value of the characteristic of the interval $(x_{i-1}, x_i)$, will be the value of the minorant at the point of its minimum at the interval $(x_{i-1},x_i), 1\leq i\leq k+p$. In accordance with Step 4 of the algorithm, a new trial is carried out in the interval with the lowest value of the minorant. And in accordance with Step 5, the point of the new trial coincides with the minorant minimum point.

@@

To solve the optimization problem, a combination of the parallel global search algorithm and a local optimization method was used. In the first stage of the search, the parallel asynchronous algorithm described in Section \ref{sec_GSA} was used. 101 processes were involved in the launch, of which one process was a master and the remaining 100 were workers (calculators). The values of the method parameters were set as follows: the maximum number of trials $K_{max}=3000$, the search accuracy $\varepsilon=10^{-3}$, the reliability parameter $r=3$, the evolvent accuracy parameter $m=12$. At the second stage, the solution was refined using the pattern search method with an accuracy of $\varepsilon=10^{-4}$ and a limit on the number of trials $K_{max}=500$.
The total time for solving the problem was 8109 seconds, an optimum point was found with the objective function value $\phi^* = 0.002115$. Thus, using the developed algorithm, the inverse problem of chemical kinetics was solved and the kinetic parameters of the reaction were found, which are given in the table \ref{tab_kinetics}.

@@

\funding{This research was funded by RFBR, project No.\,19-37-60014 (L.E.; mathematical model), and by the Russian Science Foundation, project No.\,21-11-00204 (K.B, I.L., E.K.; parallel algorithm)}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The article is devoted to the problem of developing a kinetic model of a complex chemical reaction using a parallel optimization method. The design of the kinetic model consists in finding the kinetic parameters of the reaction, which cannot be calculated analytically, and since the chemical reaction involves many stages, the optimization problem is \textcolor[rgb]{1,0,0}{multiextremal}. As a chemical reaction, the process of catalytic isomerization of the pentane-hexane fraction is considered, which is now important due to the switch of the oil refining industry to the production of gasoline corresponding to the Euro-5 standard. To solve the optimization problem, the authors have developed parallel global search algorithm and a program based on Lipschitz optimization. The kinetic parameters found made it possible to develop a mathematical model of the process, which gives good agreement with industrial data.}

@@

In the complex problems of chemical kinetics, kinetic constants of reactions cannot be found analytically, they can only be found using numerical methods (see, for example, \cite{Zaynullin2020,Enikeeva2020,Koledina2019}). In this case, the quality criterion (objective function) is set in the form of a certain algorithm, which includes the stage of numerical modeling, which, in turn, makes it compute-intensive. Furthermore, in the inverse problems of chemical kinetics, the objective function can be substantially multiextremal, i.e., having many local extremes along the global one. 
\textcolor[rgb]{1,0,0}{Gradient-based optimization methods will not give a good result for such problems. On the one hand, such methods converge only to a local solution. On the other hand, in the considered problem the gradient of the objective function is not known analytically and its numerical estimate will be time-consuming operation.}

@@

It is also possible to construct deterministic methods of global search, which are different from brute force algorithms and metaheuristic algorithms. Such methods are associated with the presence and consideration of a priori assumptions about the problem properties. 
One of the natural assumptions about the problem is that changing the values of an objective function is limited by changes of its argument. In this case, the function satisfies the Lipschitz condition, and the problem itself is called the Lipschitz global optimization problem. The central direction of the development of modern methods of Lipschitz optimization is the development of parallel algorithms. It is assumed that the values of the objective function are determined as a result of solving some other task and require a costly computational experiment.
\textcolor[rgb]{1,0,0}{On many benchmarks Lipschitz global optimization methods outperform metaheuristic algorithms in terms of the number of correctly solved problems \cite{Sergeyev2018,Sovrasov2019}.}

@@

The mathematical model involves such parameters as the raw material flow rate $F$ in the reactor and temperature $T$, but these two parameters also depend on $\tau$, therefore it is necessary to write equations on them, because technological process is non-isothermal:
\begin{linenomath}
\begin{equation}\label{eq_temper}
\dfrac{dT}{d\tau} = \dfrac{\sum_{i=1}^I\left( \dfrac{dx_i}{d\tau} \cdot \Delta H^i_{f,T} \right) }{\sum_i^J \left(x_i C^T_{p_i} \right) },
\end{equation}
\begin{equation}\label{eq_flow}
\dfrac{dF}{d\tau} = \sum_{i=1}^I  \dfrac{x_i}{d\tau}
\end{equation}
\end{linenomath}
with initial condition: at $\tau = 0$, $T(0) = T^0$, where $T$ is temperature, K, $\Delta H^i_{f,T}$ is enthalpy of the $i$-th component formation at temperature $T$, J/mol, $C^T_{p_i}$ is specific heat capacity of the $i$-th component at temperature $T$, (J/mol$\cdot$K), $F$ is mole discharge of the flow, kmol/h.

@@

\begin{linenomath}
\begin{equation}\label{FF}
FF = \dfrac{\sum_{i=1}^I \left| \dfrac{y_i^{exp} - y_i^{calc}}{y_i^{exp}} \right|}{I} \rightarrow \min,
\end{equation}
\end{linenomath}
where I is the number of observed parameters (I=19, of which 17 parameters are $x_i$, one parameter is $T$ and one parameter is $F$),  $y_{i}^{exp}$ are industrual data, $y_{i}^{calc}$ are numerical values of concentrations of reaction substances obtained by solving a direct kinetic problem. As a result of solving the inverse problem, the kinetic parameters $k^0_j$ and $E_j$ included in the Arrhenius equation \eqref{eq_rate} will be obtained. Since the number of reaction stages is 48, the kinetic model will contain 96 kinetic parameters (activation energy $E_j$ and pre-exponential multiplier $k^0_j$ for each stage), however, the authors decided to fix some of the parameters as already known for chemical reasons, so the number of optimized parameters was reduced to 24.

@@

As stated above, from a mathematical point of view, the inverse problem of chemical kinetics can be considered a Lipschitz global optimization problem.
\textcolor[rgb]{1,0,0}{This fact has has the following justification.
We solve the ODE system (\ref{eq_concentrations})--(\ref{eq_flow}) to compute the value of the objective function (\ref{FF}). The right-hand side of the ODE system is continuous functions with bounded derivatives. Thus, the system solution will also be continuous and bounded at the bounded period of time. Therefore, the objective function (\ref{FF}) will satisfy the Lipschitz condition, but the Lipschitz constant will be a priori unknown.}

In general, the problem of the specified class can be formulated as follows:
\begin{gather}
 \phi^* = \phi(y^\ast)=\min{\left\{\phi(y):y\in D\right\}}, \label{problemN}\\
 D=\left\{y\in R^N: a_i\leq y_i \leq b_i, \;  1\leq i \leq N\right\} \label{D},
\end{gather}
where $a,b \in R^N$ are boundaries of the search domain, and the objective function $\phi(y)$ satisfies the Lipschitz condition
\begin{equation}\label{Lip}
\left|\phi(y_1)-\phi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D.
\end{equation}

The criterion $\phi(y)$ is assumed to be multiextremal and is given in the form of some computational algorithm. 
\textcolor[rgb]{1,0,0}{This assumption is typical for many applied optimization problems (see, e.g., \cite{Kvasov2013,Kalyulin2017}).}
The input of this algorithm is supplied with a parameter vector, and the output is the corresponding function value. Such functions are often referred to as ``black-box'' functions. It is also assumed that each calculation of the function value at the search domain point (hereinafter also referred to as trial) is a laborious operation and requires significant computational resources. As noted in the introduction, this statement of the problem fully corresponds to the inverse problem of chemical kinetics.

@@

When describing a parallel algorithm, let us assume that there is one process master and $p$ workers. The \textcolor[rgb]{1,0,0}{superscript} will indicate the serial number of the trial performed, and the \textcolor[rgb]{1,0,0}{subscript} will be used to number the trials in ascending order of the coordinate.

@@

Step 1. Renumber \textcolor[rgb]{1,0,0}{t}he points of the set 
\[
X_k = \left\{x^1, x^2,...,x^{k+p} \right\},
\]
which contains all the points at which the trials either have been carried out or are being carried out by the lower index in ascending order, i.e.
\[
0=x_1<x_2<...<x_{x+p}=1.
\]
Step 2. Calculate values
\[
M_1=\max \left\{ \frac{ \left|z_i - z_{i-1} \right|}{(x_i-x_{i-1})^{1/N}} : x_{i-1} \notin I_k, x_i \notin I_k, 2\leq i\leq k+p \right\},
\]
\[
M_2=\max \left\{ \frac{ \left|z_{i+1} - z_{i-1} \right|}{(x_{i+1}-x_{i-1})^{1/N}} : x_i \in I_k, 2\leq i < k+p \right\},
\]
\[
M=\max\{M_1,M_2\},
\]
where $z_i=\phi(y(x_i))$, if $x_i \notin I_k, \; 1\leq i \leq k+p$.  If the value $M$ is 0, assign $M=1$. The values $z_i=\phi(y(x_i))$ at $x_i \in I_k$ are undefined because the trials at $x_i \in I_k$ are not finished yet.

@@

The master stops the algorithm if either the condition $\Delta_{t}<\varepsilon$ (\textcolor[rgb]{1,0,0}{where $t$ is the number} of the interval with the maximum characteristic from Step 4) or the condition $k>K_{max}$ (where $k$ is the number of trials performed) is met.
The real number $\varepsilon>0$ and the integer $K_{max} \geq 1$ are the algorithm parameters which correspond to the accuracy of the solution search and the maximum number of trials.

The sequential global search algorithm, on the basis of which the proposed asynchronous algorithm was implemented, allows the following interpretation \cite{Molinaro2001}. Using the points of previously performed trials, it is possible to construct a minorant of the objective function, for which the value $-R(i)$, where $R(i)$ is the value of the \textcolor[rgb]{1,0,0}{characteristics} of the interval $(x_{i-1}, x_i)$, will be the value of the minorant at the point of its minimum at the interval $(x_{i-1},x_i), 1\leq i\leq k+p$. In accordance with Step 4 of the algorithm, a new trial is carried out in the interval with the lowest value of the minorant. And in accordance with Step 5, the point of the new trial coincides with the minorant minimum point.

@@

To solve the optimization problem, a combination of the parallel global search algorithm and a local optimization method was used. In the first stage of the search, 
\textcolor[rgb]{1,0,0}{the parallel asynchronous algorithm described in Section \ref{sec_GSA} was used. The core parallel global search algorithm is available on GitHub, \url{https://github.com/UNN-ITMM-Software/gsa_nlp_solver}.}
101 processes were involved in the launch, of which one process was a master and the remaining 100 were workers (calculators). The values of the method parameters were set as follows: the maximum number of trials $K_{max}=3000$, the search accuracy $\varepsilon=10^{-3}$, the reliability parameter $r=3$, the evolvent accuracy parameter $m=12$. At the second stage, the solution was refined
\textcolor[rgb]{1,0,0}{using the Hooke-Jeeves pattern search method (see, e.g., \cite{HookJeeves})}
with an accuracy of $\varepsilon=10^{-4}$ and a limit on the number of trials $K_{max}=500$.
The total time for solving the problem was 8109 seconds, an optimum point was found with the objective function value $\phi^* = 0.002115$. Thus, using the developed algorithm, the inverse problem of chemical kinetics was solved and the kinetic parameters of the reaction were found, which are given in the table \ref{tab_kinetics}.

@@

\funding{This research was funded by RFBR, project No.\,19-37-60014 (L.E.; mathematical model), and by the Russian Science Foundation, project No.\,21-11-00204 (K.B, I.L., E.K.; parallel global optimization algorithm)}
